---
title: "FOR/STT 875, Exercise 5"
name: "Yining(Jenny) Chen"
date: "7/23/2019"
output: html_document
---

## Learning objectives
  + Practice with lists, data frames, and associated functions
  + Summarize variables held in lists and data frames
  + Work with R's linear regression `lm` function output
  + Review logical subsetting of vectors for partitioning and assigning of new values
  + Generate and visualize data from mathematical functions
  
## Overview
Modify this R Markdown file by filling in the code needed to answer the questions.

## Submission instructions
Upload your exercise_5.Rmd and exercise_5.html files to the Exercise 5 D2L dropbox. 

## Grading
You will receive full credit if your R Markdown document: 1) compiles without error; and 2) correctly completes the `# TODO:` code chunks. Also, please, fill in the feedback [questions](#questions) at the end of the exercise. 

## Getting started
Again, set some global options to make the R code printed in the HTML output look nice.

```{r global_options}
knitr::opts_chunk$set(comment = NA, tidy = TRUE)
```

## Lists

The `lm` function fits a linear model and returns a list containing elements such as the coefficients of the fitted model, the residual values, etc. Built in to R is a dataset with the fertility rate and several socioeconomic variables for 47 French-speaking provinces in Switzerland from the late 1800s. Of interest is how the socioeconomic values are related to the fertility rate. First we make the data available, and take a quick look.

```{r}
data(swiss)
str(swiss)
head(swiss)
cor(swiss)
```

From the correlation matrix we see that `Education` has the strongest correlation with `Fertility`. Next we draw a scatter plot of the two variables to assess linearity. Here are two different ways to request the scatter plot. The second matches the syntax for specifying a linear model in the `lm` function.

```{r}
plot(swiss$Education, swiss$Fertility)
plot(Fertility ~ Education, data = swiss)
```

The scatter plots don't give major causes for concern, although it looks like the outlier at the bottom right influences the correlation and the slope of a fitted least squares line substantially, but we will ignore those issues and press ahead. Next we fit a linear model, assign it to the R object `swiss.lm`, ask whether the object is a list, and obtain the names of the elements of `swiss.lm`.

```{r}
swiss.lm <- lm(Fertility ~ Education, data = swiss)
is.list(swiss.lm)
names(swiss.lm)
```

We see that `swiss.lm` is a list with 12 components. 

We can extract the components in several ways. First we will use the `$` method, familiar from working with data frames. For example `swiss.lm$residuals` returns the residuals from the model. Use this method to extract the coefficients and the residuals from the model. Assign the results to R objects called `swiss.lm.coef.1` and `swiss.lm.resid.1`. 

```{r}
swiss.lm.coef.1 <- swiss.lm$coefficients
swiss.lm.resid.1 <- swiss.lm$residuals
```

Now display the coefficients and the residuals. 
```{r}
swiss.lm.coef.1
swiss.lm.resid.1
```

We might expect that the resulting objects are vectors. Use `is.vector` to test whether this is the case.

```{r}
is.vector(swiss.lm.coef.1)
is.vector(swiss.lm.resid.1)
```

There are several other ways to extract components of a list. Here are a few ways to extract the coefficients.

1. `swiss.lm$coefficients`
2. `swiss.lm[["coefficients"]]`
3. `swiss.lm["coefficients"]`

(We could also use the number `1` in place of `"coefficients"` in (2) and (3).)

We've already used method 1 and assigned the result to `swiss.lm.coef.1`. Use the second and third methods, and assign the results to `swiss.lm.coef.2` and `swiss.lm.coef.3`. Also display all three results below.

```{r}
# TODO 5.1: Create swiss.lm.coef.2 and swiss.lm.coef.3 here, then print swiss.lm.coef.1, swiss.lm.coef.2, and swiss.lm.coef.3
swiss.lm.coef.2 <- swiss.lm[["coefficients"]]    
swiss.lm.coef.3 <- swiss.lm["coefficients"]     
# Print    
swiss.lm.coef.1      
swiss.lm.coef.2    
swiss.lm.coef.3    
```

Do the resulting objects look any different? Use `is.vector` and `is.list` on each of the three objects to determine whether they are vectors or lists.

```{r}
# TODO 5.2: Apply is.vector and is.list on swiss.lm.coef.1, swiss.lm.coef.2, and swiss.lm.coef.3 here
is.vector(swiss.lm.coef.1)      
is.vector(swiss.lm.coef.2)      
is.vector(swiss.lm.coef.3)     
is.list(swiss.lm.coef.1)   
is.list(swiss.lm.coef.2)   
is.list(swiss.lm.coef.3)    
```

You should have discovered that Methods 1 and 2 returned a vector, but Method 3 returned a list. The difference may seem subtle and not particularly important, but it will become much more important when we start to write our own functions.

## Subsetting with logical vectors

Subsetting with logical vectors is an essential skill. We'll practice this using the `gapminder` data frame. 

```{r}
gapminder <- read.delim("http://blue.for.msu.edu/FOR875/data/gapminder.tsv")
str(gapminder)
```

Here are two examples. 

First we compute the mean population of France for the years contained in the dataset.

```{r}
mean(gapminder$pop[gapminder$country == "France"])
```

Next we extract the countries in the continent `Africa`. Since the countries are repeated for each of the  years, we'll use `unique` to list the countries only once.

```{r}
unique(gapminder$country[gapminder$continent == "Africa"])
```

Next, draw a boxplot of the GDP for the year 1952.

```{r}
boxplot(gapminder$gdpPercap[gapminder$year == 1952])
hist(gapminder$gdpPercap[gapminder$year == 1952])
boxplot(gapminder$gdpPercap[gapminder$year == 2002])
hist(gapminder$gdpPercap[gapminder$year == 2002])
```

We can combine conditions with the operators `&` and `|` which represent "and" and "or" for vectors. For example, here's how to extract all the countries whose population is more than 150 million in the year 1992.

```{r}
gapminder$country[gapminder$pop > 150000000 & gapminder$year == 1992]
```

Use these ideas to extract the population of all countries whose life expectancy is more than 70 years, during 1967.

```{r}
# TODO 5.3: Add code to get the population of all countries meeting the noted conditions
gapminder$pop[gapminder$lifeExp > 70 & gapminder$year == 1967]
```

Another useful operator is `%in%`. For example here's how to compute the mean population for the United States, Mexico, and Canada during 1997.

```{r}
mean(gapminder$pop[gapminder$country %in% c("United States", "Mexico", "Canada") & gapminder$year == 1997])
```

Subscripting also can be used to change values of existing R objects as in the following example.

```{r}
dd <- data.frame(x = c("dog", "cat", "oink", "pig", "oink", "cat", "dog", "cat", "dog"), y = c("dog", "cat", "cat", "pig", "cow", "dog", "dog", "cat", "pig"), stringsAsFactors = FALSE)
dd
dd$x[dd$x == "oink"] <- "pig"
dd
dim(dd)
dd$same <- rep("no", dim(dd)[1])
dd
dd$same[dd$x == dd$y] <- "yes"
dd
```

Next we will add a variable called `G8` to the `gapminder` data frame which will be equal to 1 or 0 depending on whether the country is in the G8 group of nations, which includes France, Germany, Italy, the United Kingdom, Japan, the United States, Canada, and Russia.

First, create a variable called `G8`, all of whose values are 0, in the gapminder data frame. 

```{r}
# TODO 5.4: Add the G8 variable to the gapminder data frame here
gapminder$G8 <- rep(0, length.out = dim(gapminder)[1])    
```

Now set `G8` equal to 1 if the country is in the G8. One way to do this is to give 8 separate R commands, one for each of the G8 countries. Another is to use `%in%` and complete the work in one line.

```{r}
# TODO 5.5: Set the G8 variable to 1 for G8 countries here    
gapminder$G8[gapminder$country %in% c("France", "Germany", "Italy", "United Kingdom", "Japan", "United States", "Canada", "Russia")] <- 1    
head(gapminder)     
```

## Patterned data

We'll plot the function $f(x) = \cos(x)$ for $x$ between $-2\pi$ and $2\pi$.

First, create a vector `x` containing 1000 equally spaced values between $-2\pi$ and $2\pi$ using the `seq` function.

```{r}
##Create x here
x <- seq(from=-2*pi, to=2*pi, length.out=1000)    
```

Next use the `plot` command to plot the function. The first argument to `plot` should be the vector `x` we just created. The second argument should be `cos(x)`. 

```{r}
# TODO 5.6: Draw the plot here
x<- seq(from = -2*pi, to = 2*pi, length.out = 1000) 
plot(x, y=cos(x))
```

The graphic is OK, but since we're plotting the graph of a function, we should connect the points rather than plotting each of the points. Add the argument `type = "l"` to accomplish this.

```{r}
# TODO 5.7: Draw the second plot here    
x<- seq(from = -2*pi, to = 2*pi, length.out = 1000) 
plot(x, y=cos(x), type = "l")   
```

As seen in the G8 example above, `rep` is another very useful function for generating patterned data. The `rep` function replicates its first argument. Important arguments include `length.out`, `times`, and `each`. Some examples:

```{r}
ltrs <- c("a", "b", "q", "z")
rep(ltrs, length.out = 9)
rep(ltrs, each = 3)
rep(ltrs, times = 3)
rep(ltrs, times = c(3, 2, 4, 7))
```

********

Congratulations! You've reached the end of Exercise 5.

###Questions?

If you have any lingering questions about the material in this document or in the corresponding chapter, put them here.

*None*


###Give us your feedback!

1.  How do you feel you're doing in this class?

*Piazza is really helpful. Found out all my questions had been asked by other classmates and had the answers already.*

2.  What could be done to improve your learning experience?

*None. Perfect!*

*********
